---
title: 'linear_interpolation example'
author: "Bill Gillespie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: default
  pdf_document: default
sponsor: MetrumRG
projectnumber: Torsten
---

# Toy example illustrating use of the linear_interpolation function

## Setup R environment

```{r setup}

rm(list = ls())
gc()

modelName <- "testInterp2"

## Relative paths assuming the working directory is the script directory
## containing this script
scriptDir <- getwd()
toolsDir <- file.path(scriptDir, "tools")
tempDir <- file.path(scriptDir, "temp")
dir.create(tempDir)

##.libPaths("lib")

## Path for cmdstan interface to Stan augmented with the Torsten 
## function library.
stanDir <- file.path(dirname(scriptDir), "cmdstan")

library(rstan)
library(bayesplot)
library(tidyverse)
library(parallel)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

set.seed(10271998) ## not required but assures repeatable results

## A couple handy-dabdy functions

mcmcHistory <- function(fit, pars = names(fit), nParPerPage = 6, myTheme = NULL){
    require(tidyverse)
    require(bayesplot)
    posterior <- as.array(fit, pars = pars)
    pars <- dimnames(posterior)[[3]]
    pnuts <- nuts_params(fit)

    nPars <- length(pars)
    nPages <- ceiling(nPars / nParPerPage)
    parameters <- data.frame(parameter = pars,
                             page = sort(rep(1:nPages, length = nPars)),
                             stringsAsFactors = FALSE)

    for(i in 1:nPages){
      posterior <- as.array(fit, pars = with(parameters, pars[page == i]))
      if(sum((pnuts %>% filter(Parameter == "divergent__"))$Value)){
            print(mcmc_trace(posterior,
                             divergences = pnuts,
                             facet_args = list(ncol = 1, strip.position = "left")) +
                  myTheme +
                  scale_x_continuous(breaks = seq(0, nPost, len = 5)))
            
        }else{
            print(mcmc_trace(posterior,
                             facet_args = list(ncol = 1, strip.position = "left")) +
                  myTheme +
                  scale_x_continuous(breaks = seq(0, nPost, len = 5)))
        }
    }
    NULL
}

mcmcDensity <- function(fit, pars = names(fit), byChain = FALSE, nParPerPage = 16, 
                        myTheme = NULL, prior = NULL){
  require(tidyverse)
  require(bayesplot)
  posterior <- as.array(fit, pars = pars)
  pars <- dimnames(posterior)[[3]]
  pnuts <- nuts_params(fit)
  
  nPars <- length(pars)
  nPages <- ceiling(nPars / nParPerPage)
  parameters <- data.frame(Parameter = pars,
                           page = sort(rep(1:nPages, length = nPars)),
                           stringsAsFactors = FALSE)
  
  if(!is.null(prior)) prior <- prior %>% left_join(parameters)

  for(i in 1:nPages){
    posterior <- as.array(fit, pars = with(parameters, pars[page == i]))
    if(byChain){
      p1 <- mcmc_dens_overlay(posterior)
    }else{
      p1 <- mcmc_dens(posterior)
    }
    if(!is.null(prior))
      p1 <- p1 + geom_line(data = subset(prior, page == i), 
                           aes(x = value, y = density),
                           color = "red")
    print(p1 + myTheme)
  }
  NULL
}

compileModel <- function(model, stanDir = stanDir){
    modelName <- basename(model)
    dir.create(model)
    file.copy(paste(model, "stan", sep = "."), file.path(model, paste(modelName, "stan", sep = ".")),
              overwrite = TRUE)
    model <- file.path(model, modelName)
    system(paste("make --directory=", stanDir, " ", model, sep = ""))
}

runModel <- function(model, data, iter, warmup, thin, init, seed, chain = 1,
                     stepsize = NULL, adapt_delta = NULL, refresh = 1,
                     save_warmup = 0, algorithm = "hmc", engine = "nuts"){
  modelName <- basename(model)
  model <- file.path(model, modelName)
  system(paste(model, " sample",
               " algorithm=", algorithm,
               ifelse(is.null(engine), "", paste(" engine=", engine, sep = "")),
               ifelse(is.null(stepsize), "", paste(" stepsize=", stepsize, sep = "")),
               " num_samples=", iter - warmup,
               " num_warmup=", warmup,
               " save_warmup=", save_warmup,
               " thin=",  thin,
               ifelse(is.null(adapt_delta), "", paste(" adapt delta=", adapt_delta, sep = "")),
               " data file=", data,
               ifelse(is.null(init), "", paste(" init=", init, sep = "")),
               " random seed=", seed,
               " output file=", paste(model, chain, ".csv", sep = ""),
               " refresh=", refresh,
               sep = ""))
}

```

## Generate data set and initial estimates

Generate data from a parabolic function on equally spaced points on the x axis.

```{r data}

nObs <- 23
xObs <- seq(0, 10, length = nObs)
yObs <- rnorm(nObs, (xObs - 5)^2, 2)
nPred <- 51
xmin <- 0
xmax <- 10
dx <- (xmax - xmin) / (nPred - 1)
xPred <- xmin + (1:nPred - 1) * dx
nx <- 5

## create data set
data <- list(nObs = nObs,
             xObs = xObs,
             yObs = yObs,
             nx = nx,
             nPred = 51,
             xPred = xPred)

### create initial estimates
init <- function(){
    list(y = rnorm(nx, 10, 5),
         xSimplex = rep(1 / (nx - 1), nx - 1),
        sigma = 5)
}

```

## Stan model

```{r stan model}

stan_model <- "
data{
  int nObs;
  real xObs[nObs];
  real yObs[nObs];
  int nx;
  int nPred;
  real xPred[nPred];
}

transformed data{
  real xmin = min(xObs);
  real xmax = max(xObs);
}

parameters{
  real y[nx];
  real<lower = 0> sigma;
  simplex[nx - 1] xSimplex;
}

transformed parameters{
  real yHat[nObs];
  real x[nx];

  x[1] = xmin;
  x[nx] = xmax;
  for(i in 2:(nx-1))
    x[i] = x[i-1] + xSimplex[i-1] * (xmax - xmin);

  yHat = linear_interpolation(xObs, x, y);
}

model{
  xSimplex ~ dirichlet(rep_vector(1, nx - 1));
  y ~ normal(0, 25);
  yObs ~ normal(yHat, sigma);
}

generated quantities{
  real yHatPred[nPred];
  real yPred[nPred];

  yHatPred = linear_interpolation(xPred, x, y);
  for(i in 1:nPred)
    yPred[i] = normal_rng(yHatPred[i], sigma);

}"

```

## Run the Stan model

We are using the cmdstan interface to Stan augmented with the Torsten function library.

```{r run Stan}

## Specify the variables for which you want history and density plots

parametersToPlot <- c("x", "y", "sigma")

## Additional variables to monitor
otherRVs <- c("yPred")

parameters <- c(parametersToPlot, otherRVs)

## run Stan

nChains <- 4
nPost <- 500 ## Number of post-burn-in samples per chain after thinning
nBurn <- 500 ## Number of burn-in samples per chain after thinning
nThin <- 10

nIter <- (nPost + nBurn) * nThin
nBurnin <- nBurn * nThin

RNGkind("L'Ecuyer-CMRG")
mc.reset.stream()

write(stan_model, file = file.path(tempDir, paste(modelName, "stan", sep = ".")))
compileModel(model = file.path(tempDir, modelName), stanDir = stanDir)

chains <- 1:nChains

mclapply(chains,
         function(chain, model, data, iter, warmup, thin, init, 
                  tempDir = "temp"){
             tempDir <- file.path(tempDir, chain)
             dir.create(tempDir)
             with(data, stan_rdump(ls(data), 
                                   file = file.path(tempDir, "data.R")))
             inits <- init()
             with(inits, stan_rdump(ls(inits), 
                                    file = file.path(tempDir, "init.R")))
             runModel(model = model, data = file.path(tempDir, "data.R"),
                      iter = iter, warmup = warmup, thin = thin,
                      init = file.path(tempDir, "init.R"), 
                      seed = sample(1:999999, 1),
                      chain = chain, refresh = 10,
                      adapt_delta = 0.95, stepsize = 0.01)
            },
         model = file.path(tempDir, modelName),
         data = data,
         init = init,
         iter = nIter, warmup = nBurnin, thin = nThin,
         mc.cores = min(nChains, detectCores()))

fit <- read_stan_csv(file.path(tempDir, modelName, paste(modelName, chains, ".csv", sep = "")))

```

## Convergence diagnostics and parameter summaries

```{r parameters}

options(bayesplot.base_size = 12,
        bayesplot.base_family = "sans")
color_scheme_set(scheme = "brightblue")
myTheme <- theme(text = element_text(size = 12), axis.text = element_text(size = 12))

rhats <- rhat(fit, pars = parametersToPlot)
mcmc_rhat(rhats) + yaxis_text() + myTheme

ratios1 <- neff_ratio(fit, pars = parametersToPlot)
mcmc_neff(ratios1) + yaxis_text() + myTheme

mcmcHistory(fit, pars = parametersToPlot, nParPerPage = 5, myTheme = myTheme)
mcmcDensity(fit, pars = parametersToPlot, nParPerPage = 16, byChain = TRUE, 
            myTheme = theme(text = element_text(size = 12), axis.text = element_text(size = 10)))
mcmcDensity(fit, pars = parametersToPlot, nParPerPage = 16, 
            myTheme = theme(text = element_text(size = 12), axis.text = element_text(size = 10)))

pairs(fit, pars = parametersToPlot[!grepl("rho", parametersToPlot)])

print(fit, pars = parametersToPlot)

```

## Posterior predictive distributions

```{r PPC}

xdata <- with(data,
              data.frame(x = xObs, y = yObs))
xPred <- with(data,
              data.frame(x = xPred))

pred <- as.data.frame(fit, pars = "yPred") %>%
  gather(factor_key = TRUE) %>%
  group_by(key) %>%
  summarize(lb = quantile(value, probs = 0.05, na.rm = TRUE),
            median = quantile(value, probs = 0.5, na.rm = TRUE),
            ub = quantile(value, probs = 0.95, na.rm = TRUE)) %>%
  bind_cols(xPred)

p1 <- ggplot(xdata, aes(x = x, y = y))
p1 <- p1 + geom_point() +
  labs(x = "x",
       y = "y") +
  theme(text = element_text(size = 12), axis.text = element_text(size = 12),
        legend.position = "none", strip.text = element_text(size = 8))

print(p1 + geom_line(data = pred, aes(x = x, y = median)) +
        geom_ribbon(data = pred, aes(x = x, ymin = lb, ymax = ub), inherit.aes = FALSE, alpha = 0.25))

```
